Algorithm explanation

### 2013-09-25 article/experiment.tex
Minor fix
Amendment

We have incorporated our algorithm to the software model checker for
higher-order programs (MoCHi) \cite{conf/pldi/KobayashiSU11}.  It is
specifically designed to verify the correctness of higher-order
programs written in a small subset of ML.  This verification tool
abstracts a given input program by inferring dependent types for
higher-order programs \cite{conf/ppdp/UnnoK09}.

We implemented our proposed algorithm into MoCHi by two steps.  First,
we replace an existing theorem prover with our interpolating theorem
prover as a sub-routine for solving Horn clauses in dependent typing.
We used a method to obtain a common small interpolant across different
interpolating problems, while reducing a Horn clause problem into
interpolation \cite{conf/ppdp/UnnoK09}.  Second, we completely
replaced MoCHi's Horn solving algorithm with our algorithm.

@@

There are some interpolating theorem provers known.
\begin{itemize}
\item \textsc{CSIsat} \cite{conf/cav/BeyerZM08} combines linear
  arithmetic reasoning and SAT-based logical reasoning.
\item \textsc{FOCI} \cite{website/foci} extracts an interpolant by the
  proof-based method.
\item \textsc{CLP-Prover} \cite{website/clp} generates an interpolant
 from constraints by linear arithmetic reasoning. It only supports
 conjunctive problems.
\end{itemize}

In our experiment, we compared the performance of our algorithm with
CSIsat, that is mainly used as a sub procedure of MoCHi.  Note that
CSIsat may return different interpolants for an interpolating problem
because of the randomized algorithm.

\paragraph{Comparison over simple interpolation}
First, we compare by simply generating an interpolant without
incorporating into MoCHi.  \textsc{Yint} represents our interpolating
theorem prover.

\begin{itemize}
\item Disjunctive interpolation in $A$.
\[\begin{array}{r l}
\varphi_A & (x-1 \geq 0) \vee (x+1 \leq 0) \\
\hline
\varphi_B & x=0
\end{array}\]
The result for this:
\[\begin{array}{l c}
\text{\textsc{Yint}} & (x+1 \leq 0) \vee (-x+1 \leq 0) \\
\text{\textsc{CSIsat}} & (-x \leq -1 \vee x \leq -1)
\end{array}\]
\item Disjunctive interpolation in $B$.
\[\begin{array}{r l}
\varphi_A & x=0 \\
\hline
\varphi_B & (x-1 \geq 0) \vee (x+1 \leq 0)
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & -x=0 \\
\text{\textsc{CSIsat}} & (-x \leq 0 \wedge x \leq 0)
\end{array}\]
\item Complex formula in $A$.
% complicated disj
\[\begin{array}{r l}
\varphi_A & ((x-5<0) \wedge (-x<0)) \vee ((x<0) \wedge (-x-5<0)) \\
\hline
\varphi_B & x=0
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & (x+1 \leq 0) \vee (-x+1 \leq 0) \\
\text{\textsc{CSIsat}} & (x < 0 \vee -x < 0)
\end{array}\]
\end{itemize}
For those problems that are simple enough, Yint and CSIsat returns the
same results.

Next we tried the problem that an equality interpolant exists.
% equality
\[\begin{array}{r l}
\varphi_A & x=0 \\
\hline
\varphi_B & x-1=0
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & x=0 \\
\text{\textsc{CSIsat}} & x \leq 0
\end{array}\]

@@

% quantifier
\[\begin{array}{r l}
\varphi_A & (y-z=0) \wedge (x-z=0) \\
\hline
\varphi_B & (-w+y+1 \leq 0) \wedge (-w+x \geq 0)
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & x-y=0 \\
\text{\textsc{CSIsat}} & x-y \leq 0
\end{array}\]
Because our algorithm handles linear equalities special, our algorithm
makes an effort to return an interpolant of linear equality if one
exists.

Additionally, as the follwing results of \textsc{Yint} shows, if a
simple interpolant exists for a given problem, our algorithm tries
such result.
\[\begin{array}{r l}
\varphi_A & ((-y+1 \leq 0) \wedge (-x \leq 0)) \vee ((-y \leq 0) \wedge (-x+1 \leq 0)) \\
\hline
\varphi_B & ((y+1<0) \wedge (x<0)) \vee ((y<0) \wedge (x+1<0))
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & -x \leq 0 \\
\text{\textsc{CSIsat}} & (((x < 0 \vee -x \leq 0) \wedge 0 \leq x) \vee ((y < 0 \vee -y \leq 0) \wedge 0 \leq y))
\end{array}\]

@@

\[\begin{array}{r l}
\varphi_A & (x-1=0) \vee (x+1=0) \\
\hline
\varphi_B & (x-2=0) \vee (x=0) \vee (x+2=0)
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & (-x-1=0) \vee (-x+1=0) \\
\text{\textsc{CSIsat}} & ((-x \leq -1 \vee x \leq -1) \wedge (-x \leq 1 \vee - x \leq - 1) \wedge (x \leq - 1 \vee x \leq 1))
\end{array}\]

@@

\[\begin{array}{r l}
\multirow{2}{*}{$\varphi_A$}
& ((x-y-1 \leq 0) \wedge (-x+y-2 \leq 0) \wedge (x+y-1 \leq 0) \wedge (-x-y-2 \leq 0)) \vee \\
& ((x-y-2 \leq 0) \wedge (-x+y-1 \leq 0) \wedge (x+y-2 \leq 0) \wedge (-x-y-1 \leq 0)) \\
\hline
\multirow{2}{*}{$\varphi_B$}
& ((x-y+1<0) \wedge (-x+y-2 \leq 0) \wedge (x+y-2 \leq 0) \wedge (-x-y+1<0)) \vee \\
& ((x-y-2 \leq 0) \wedge (-x+y+1<0) \wedge (x+y+1<0) \wedge (-x-y-2 \leq 0))
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & (-x+17y-33 \leq 0) \wedge (x-17y-33 \leq 0) \\
\text{\textsc{CSIsat}} & ((-1 \leq (x + y) \vee (x + -y) \leq 1) \wedge (-1 \leq (x + -y) \vee (x + y) \leq 1))
\end{array}\]

@@

\[\begin{array}{r l}
\multirow{2}{*}{$\varphi_A$}
& ((y+1=0) \wedge (x-1=0)) \vee ((y+1=0) \wedge (x+1=0)) \vee \\
& ((y-1=0) \wedge (x+1=0)) \vee ((y-1=0) \wedge (x-1=0)) \\
\hline
\multirow{2}{*}{$\varphi_B$}
& ((y+2 \leq 0) \wedge (x-2 \geq 0)) \vee ((y+2 \leq 0) \wedge (x+2 \leq 0)) \vee \\
& ((y-2 \geq 0) \wedge (x+2 \leq 0)) \vee ((y-2 \geq 0) \wedge (x-2 \geq 0))
\end{array}\]
\begin{tabular}{l c}
\textsc{Yint} & $(-2y-3 \leq 0) \wedge (2y-3 \leq 0)$
\end{tabular}

@@

\[\begin{array}{r l}
\varphi_A & (x=1 \wedge y=1) \vee (x=0 \wedge y=0) \\
\hline
\varphi_B & (x=1 \wedge y \neq 1) \vee (x=0 \wedge y \neq 0)
\end{array}\]
\begin{tabular}{l c}
\textsc{Yint} & $x-y=0$
\end{tabular}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=.8]{figures/chart.png}
\end{center}
\caption{Result obtained from MoCHi with Horn clause solver CSIsat and
Yhorn}
\end{figure}

We obtained better result in some cases compared to the one we had in CSIsat.

\begin{figure}
\begin{center}
\includegraphics[scale=.5]{figures/time.eps}
\end{center}
\caption{Time consumption ration of Yhorn compared to CSIsat}
\end{figure}

>>>>

We have incorporated our algorithm to the software model checker for
higher-order programs \textsc{MoCHi} \cite{conf/pldi/KobayashiSU11}.  It is
specifically designed to verify the correctness of higher-order
programs written in a small subset of ML.  This verification tool
abstracts a given input program by inferring dependent types for
higher-order programs \cite{conf/ppdp/UnnoK09}.

We implemented our proposed algorithm into \textsc{MoCHi} by two steps.  First,
we replace an existing theorem prover with our interpolating theorem
prover as a sub-routine for solving Horn clauses in dependent typing.
We used a method to obtain a common small interpolant across different
interpolating problems, while reducing a Horn clause problem into
interpolation \cite{conf/ppdp/UnnoK09}.  Second, we completely
replaced MoCHi's Horn solving algorithm with our algorithm.

@@

As an experiment of our interpolating algorithm, we compared the
quality of resulting interpolants with
ones that \testsc{CSIsat} \cite{conf/cav/BeyerZM08} generates.
It combines linear arithmetic reasoning and SAT-based logical reasoning.
\textsc{CSIsat} is mainly used as a sub procedure of current \textsc{MoCHi}.  Note that
it may return different interpolants for an interpolating problem
because of the randomized algorithm.  Results shown below
are ones that are randomly chosen.
\textsc{Yint} represents our interpolating algorithm.

\begin{itemize}
\item Disjunctive interpolation in $A$.
\[\begin{array}{r c}
\varphi_A : & (x-1 \geq 0) \vee (x+1 \leq 0) \\
\hline
\varphi_B : & x=0
\end{array}\]
The result for this:
\[\begin{array}{l c}
\text{\textsc{Yint}} & (x+1 \leq 0) \vee (-x+1 \leq 0) \\
\text{\textsc{CSIsat}} & (-x \leq -1 \vee x \leq -1)
\end{array}\]
\item Disjunctive interpolation in $B$.
\[\begin{array}{r c}
\varphi_A : & x=0 \\
\hline
\varphi_B : & (x-1 \geq 0) \vee (x+1 \leq 0)
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & -x=0 \\
\text{\textsc{CSIsat}} & (-x \leq 0 \wedge x \leq 0)
\end{array}\]
\item Complex formula in $A$.
% complicated disj
\[\begin{array}{r c}
\varphi_A : & ((x-5<0) \wedge (-x<0)) \vee ((x<0) \wedge (-x-5<0)) \\
\hline
\varphi_B : & x=0
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & (x+1 \leq 0) \vee (-x+1 \leq 0) \\
\text{\textsc{CSIsat}} & (x < 0 \vee -x < 0)
\end{array}\]
\end{itemize}
For those problems that are simple enough, \textsc{Yint} and \textsc{CSIsat} returns the
same results.

Next we tried the problem that an equality interpolant exists.
% equality
\[\begin{array}{r c}
\varphi_A : & x=0 \\
\hline
\varphi_B : & x-1=0
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & x=0 \\
\text{\textsc{CSIsat}} & x \leq 0
\end{array}\]

@@

% quantifier
\[\begin{array}{r c}
\varphi_A : & (y-z=0) \wedge (x-z=0) \\
\hline
\varphi_B : & (-w+y+1 \leq 0) \wedge (-w+x \geq 0)
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & x-y=0 \\
\text{\textsc{CSIsat}} & x-y \leq 0
\end{array}\]
Because our algorithm handles linear equalities special, our algorithm
makes an effort to return an interpolant of linear equality if one
exists.

Additionally, as the follwing results of \textsc{Yint} shows, if a
simple interpolant exists for a given problem, our algorithm tries
such result.
\[\begin{array}{r c}
\varphi_A : & ((-y+1 \leq 0) \wedge (-x \leq 0)) \vee ((-y \leq 0) \wedge (-x+1 \leq 0)) \\
\hline
\varphi_B : & ((y+1<0) \wedge (x<0)) \vee ((y<0) \wedge (x+1<0))
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & -x \leq 0 \\
\text{\textsc{CSIsat}} & (((x < 0 \vee -x \leq 0) \wedge 0 \leq x) \vee ((y < 0 \vee -y \leq 0) \wedge 0 \leq y))
\end{array}\]

@@

\[\begin{array}{r c}
\varphi_A : & (x-1=0) \vee (x+1=0) \\
\hline
\varphi_B : & (x-2=0) \vee (x=0) \vee (x+2=0)
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & (-x-1=0) \vee (-x+1=0) \\
\text{\textsc{CSIsat}} & ((-x \leq -1 \vee x \leq -1) \wedge (-x \leq 1 \vee - x \leq - 1) \wedge (x \leq - 1 \vee x \leq 1))
\end{array}\]

@@

\[\begin{array}{r c}
\multirow{2}{*}{$\varphi_A$}
& ((x-y-1 \leq 0) \wedge (-x+y-2 \leq 0) \wedge (x+y-1 \leq 0) \wedge (-x-y-2 \leq 0)) \vee \\
& ((x-y-2 \leq 0) \wedge (-x+y-1 \leq 0) \wedge (x+y-2 \leq 0) \wedge (-x-y-1 \leq 0)) \\
\hline
\varphi_B : & (x=0 \wedge y=2) \vee (x=0 \wedge y=-2)
\end{array}\]
\[\begin{array}{l c}
\text{\textsc{Yint}} & (-x+17y-33 \leq 0) \wedge (x-17y-33 \leq 0) \\
\text{\textsc{CSIsat}} & ((-1 \leq (x + y) \vee (x + -y) \leq 1) \wedge (-1 \leq (x + -y) \vee (x + y) \leq 1))
\end{array}\]

@@

\[\begin{array}{r c}
\multirow{2}{*}{$\varphi_A$}
& ((y+1=0) \wedge (x-1=0)) \vee ((y+1=0) \wedge (x+1=0)) \vee \\
& ((y-1=0) \wedge (x+1=0)) \vee ((y-1=0) \wedge (x-1=0)) \\
\hline
\multirow{2}{*}{$\varphi_B$}
& ((y+2 \leq 0) \wedge (x-2 \geq 0)) \vee ((y+2 \leq 0) \wedge (x+2 \leq 0)) \vee \\
& ((y-2 \geq 0) \wedge (x+2 \leq 0)) \vee ((y-2 \geq 0) \wedge (x-2 \geq 0))
\end{array}\]
\begin{tabular}{l c}
\textsc{Yint} & $(-2y-3 \leq 0) \wedge (2y-3 \leq 0)$
\end{tabular}

@@

\[\begin{array}{r c}
\varphi_A & (x=1 \wedge y=1) \vee (x=0 \wedge y=0) \\
\hline
\varphi_B & (x=1 \wedge y \neq 1) \vee (x=0 \wedge y \neq 0)
\end{array}\]
\begin{tabular}{l c}
\textsc{Yint} & $x-y=0$
\end{tabular}

@@

\begin{figure}
\begin{center}
\includegraphics[scale=.6]{figures/chart.png}
\end{center}
\caption{Result obtained from \textsc{MoCHi} with Horn clause solver \textsc{CSIsat} and
\textsc{Yhorn}}
\end{figure}

We obtained better result in some cases compared to the one we had in \textsc{CSIsat}.

\begin{figure}
\begin{center}
\includegraphics[scale=.8]{figures/time.eps}
\end{center}
\caption{Time consumption ration of Yhorn compared to \textsc{CSIsat}}
\end{figure}

### 2013-09-25 article/future.tex
Small fixes
Amendment

Our proposed algorithms have still numbers of rooms for further
improvement.  Based on the experiment result from the previous
chapter, here we give raise some analysis and discussion for them.

\paragraph{Space explosion for constraint duplication}
Our experimental implementation did not simplify the linear constraint
for every predicate variable and it caused a linear constraint
unsolvable due to the shortage of space resources by exponential
growth of constraints.  As we described in the algorithm explanation,
it may be possible to perform quantifier elimination for reduce the
size of linear constraints.  On the other hand without the
optimization, the constraint duplication may occur every time a
constraint is generated for the predicate vertices with multiple
successors.

We think that the algorithm with such optimization would not cause a
memory shortage because constraints gently grow according to roughly
the number of vertices.  Although the worst case can be produced by
causing a conjunction split frequently, we consider that it is not the
common case in program verification because of our hypothesis.

In practice, the size of constraint may be limited by simplification,
and additionally, normally the algorithm does not need to preserve the
constraint until the time to build a concrete solution in the end.

\paragraph{Quantifier elimination}
It may be useful to prepare a customized algorithm for performing
quantifier elimination for linear expressions because linear
constraints that are generated from our algorithm have certain pattern
in their format.

In the preliminary experiment before incorporating into MoCHi, although
we have tried the quantifier elimination over integer theory, the
result was unsatisfactory in the sense that simplified formulas had
disjunctions.  It caused the loss of efficiency in solving constraints
and generating unsatisfiable cores.

We have also tried the quantifier elimination over real space, and the
result was better than the integer version.  However, because our
algorithm first treats all linear inequalities over integer space,
more study on the gap between these two are required when we consider
the completeness.

>>>>

Our proposed algorithms have still a number of rooms for further
improvement.  Based on the experiment result from the previous
chapter, here we give raise some analysis and discussion for them.

### 2013-09-25 article/interpolation.tex
Amendment

Our algorithm handles a linear expression $e$ which is in a form:
\[ a_1 x_1 + \cdots + a_n x_n + b \leq 0 \]
This expression contains coefficient parameters $a_1, \ldots, a_n$ and
a constant parameter $b$.  We treat a special expressions $\top$ as $0
\leq 0$ and $\bot$ as $1 \leq 0$.  The algorithm input and output is a
formula $\psi$ that consists of conjunctions and/or disjunctions of
linear expressions.
\[ \psi = e | e \wedge e | e \vee e | \top | \bot \]

In the simple interpolation, the algorithm receives two formulas in
$\psi$ and returns an interpolant in $\psi$.  For the later extension
to handle a symmetric interpolation problem, the algorithm receives
multiple formulas in $\psi$ and returns intermediate interpolants in
$\psi$.  For computing those interpolants over linear arithmetics, we
use Farkas's Lemma.

\paragraph{Farkas's Lemma on linear inequalities}
Let a linear inequality $e_i$ be represented as
$a_{i,1} x_1 + \cdots + a_{i,m} x_m <= a_{i,0}$.  Assuming that
$e_1,\cdots,e_n$ implies $e_0$, there exists
$\lambda_1,\cdots,\lambda_n$ that satisfy
\[a_{0,j} = \sum_{i=1}^n \lambda_i * a_{i,j} \quad (j=0, \ldots,m)\]

@@

Our algorithm aims to construct a relatively small interpolant in a
reasonable computation time for any input.  It also aims to obtain a
common solution among multiple interpolating problems.  These goals
are set because the algorithm is designed to be used as a sub-routine
for solving Horn clause satisfying problems
that are made while discovering predicates for abstraction
during the program verification.  These advantages of our algorithm
make it possible to return a relatively small predicates for those
problems. To accomplish these aims, the algorithm preserves a set of
interpolants during the computation, and executes operations
over interpolant sets.

\paragraph{Interpolation between conjunctions}
We first present an interpolating algorithm between two conjunctive
sets of linear inequalities $A = \left\lbrace e_1,\ldots,e_m
\right\rbrace$ and $B = \left\lbrace e_{m+1},\ldots,e_{m+n}
\right\rbrace$, where $e_i$ takes a form $\mathbf{a}_i \mathbf{x} + b
\leq 0$.  We assume that conjunctions of $A$ and $B$ are inconsistent.
For later convenience, we call this interpolating problem $\left( A, B
\right)$.  Then there exists an interpolating linear expression
$e_\star$ which satisfies
\begin{align*}
\left\lbrace e_1,\ldots,e_m \right\rbrace & \vdash e_\star \\
e_\star \wedge \left\lbrace e_{m+1},\ldots,e_{m+n} \right\rbrace & \vdash \bot
\end{align*}
For convenience, we assume that any expression $e_i$ have variables
$x_1, \ldots, x_a$.

@@

\paragraph{Symmetric Interpolation}
TODO

>>>>

Our algorithm handles a linear expression $e$ which is in a form:
\[ a_1 x_1 + \cdots + a_n x_n + b \leq 0 \]
This expression contains coefficient parameters $a_1, \ldots, a_n$ and
a constant parameter $b$.  We treat special expressions $\top$ as $0
\leq 0$ and $\bot$ as $1 \leq 0$.  The algorithm input and output is a
formula $\psi$ that consists of conjunctions and/or disjunctions of
linear expressions.
\[ \psi = e | e \wedge e | e \vee e | \top | \bot \]

In the simple interpolation, the algorithm receives two formulas in
$\psi$ and returns an interpolant in $\psi$.

\paragraph{Farkas's Lemma on linear inequalities}
Let a linear inequality $e_i$ be represented as
$a_{i,1} x_1 + \cdots + a_{i,m} x_m <= a_{i,0}$.  Assuming that
$e_1,\cdots,e_n$ implies $e_0$, there exists
$\lambda_1,\cdots,\lambda_n$ that satisfy
\[a_{0,j} = \sum_{i=1}^n \lambda_i a_{i,j} \quad (j=0, \ldots,m)\]

@@

Our algorithm aims to construct a relatively small interpolant in a
reasonable computation time for any input.  It also aims to obtain a
common solution among multiple interpolating problems.
To accomplish these goals, the algorithm preserves a set of solution
interpolants instead of one interpolant during the computation of sub
problems, and executes operations over interpolant sets.

\paragraph{Interpolation between conjunctions}
We first present an interpolating algorithm between two conjunctive
sets of linear inequalities $A = \left\lbrace e_1,\ldots,e_m
\right\rbrace$ and $B = \left\lbrace e_{m+1},\ldots,e_{m+n}
\right\rbrace$, where $e_i$ takes a form $\mathbf{a}_i \mathbf{x} + b
\leq 0$.  We assume that conjunctions of $A$ and $B$ are inconsistent.
For later convenience, we call this interpolating problem $\left( A, B
\right)$.  Then there exists an interpolating linear expression
$e_\star$ which satisfies
\begin{align*}
\left\lbrace e_1,\ldots,e_m \right\rbrace & \vdash e_\star \\
e_\star \wedge \left\lbrace e_{m+1},\ldots,e_{m+n} \right\rbrace & \vdash \bot
\end{align*}

@@
